From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-02 paper/paper.Rmd
intro to 'creator' section, working on intro/rewrite to 'informer'

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level.  
The algorithm is merely a simple and direct implementation of the master
equation, progressing in random step sizes determined
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free[^2] stochastic ecological models.

[^2]: that is, free from the approximation made by SDE models as we see in the
van Kampen example.  All models are, of course, only approximations.

@@

Stochasticity can give rise to a wide array of interesting phenomena
that are often at odds with our intuition. As these examples will
illustrate, such behavior arises through an interaction between
stochastic and non-linear elements of a system, underscoring the
importance of explicitly including stochasticity in ecological
models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
does not have to be large to give rise to these phenomena.  In this
section, I highlight several examples that have shown up frequently
in the literature and prove sufficiently tractable to provide a
simple intuition for the mechanisms involved.

@@

<!-- More frank introduction to this section, see reply to Reviewer 1-->

One of the most interesting aspects of noise is as a source of
additional information.  This can be realized in a variety of ways.
For instance, in deriving expressions for the variance created by
demographic noise above, we observed that different partitions of
a logistic growth model can lead to the same macroscopic dynamics
(same values for $b$ - $d$), while having different variance (due
to different values for $b$+$d$).  This suggests that the variation
observed in such a population could potentially help resolve
how dynamics are partitioned between individual birth and death
rates (though in practice one would have to first account for any
contribution of environmental noise).

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.  This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure \ref{tipping}.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure \ref{tipping}.

@@

>>>>

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level.  
The algorithm is merely a simple and direct implementation of the master
equation, progressing in random step sizes determined
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free^[that is, free from the approximation
made by SDE models as we see in the van Kampen example.  All models are, of course,
only approximations.] stochastic ecological models.

@@

Outside of the primarily theoretical literature on stochastic dynamics
in ecology, noise is viewed largely as a nusiance: something entirely
distinct from the processes we are interested in which invariably keeps
getting in the way.  As we have seen in section on origins, such 
characterizations are misleading: noise arises from the very same processes.
While both empirical and statistical literature have frequently emphasized
techniques to remove noise, a central theme of stochastic processes
have emphasized an almost opposite conclusion: noise cannot be subtracted or
averaged away, since tthe presence of noise can fundamentally alter the
dynamics we observe.
Stochasticity can give rise to a wide array of interesting phenomena
that are often at odds with our intuition. As these examples will
illustrate, such behavior arises through an interaction between
stochastic and non-linear elements of a system, underscoring the
importance of explicitly including stochasticity in ecological
models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
does not have to be large to give rise to these phenomena.  In this
section, I highlight several examples that have shown up frequently
in the literature and prove sufficiently tractable to provide a
simple intuition for the mechanisms involved.

@@

In examining "noise the creator," we began with a tension between
the intepretation of noise as it is usually presented in the context of
statistical inference: a nusiance to be stripped away, and how it is
viewed by theorist, as a potentially fundamental driver of patterns,
including oscillations, regime shifts, extinction & coexistence.
In this third section, we attempt to re-unite these two largely
divergent views of the role of noise by highlighting how phenomena
from the latter (noisy phenomena) can be used to address the questions
of the former (inferring underlying process from pattern).

The most provacative example of how noisy phenomena can act to reveal
underlying processes comes from the literature on so-called 'early
warning signs' of critical transitions [e.g. @Scheffer2009].  This approach
exploits the ability of the equations we encountered in "origins of noise"
to reveal subtle changes in underlying processes before those changes are
visible at the macroscopic (deterministic) level.  This literature focuses
on the phenomenon of a tipping point, in which gradual changes to the environment
show essentially no changes to the behavior of a system (i.e. the abundance of a
population or health of lake ecosystem) followed by a rapid change when the
tipping point is reached that cannot be easily reversed. [@Scheffer2009;
@Scheffer2015].  The fundamental insight of early warning signals has been
to realize that though the average or deterministic dynamics of such a system
do not change significantly prior to the tipping point, the dynamics of the
noise must change in a consistent and predictable manner.

This pattern, known as 'critical slowing down' for reasons that will become 
obvious, was first proposed an an early warning sign for ecological
transitions by @Wissel1984, but has received considerable attention only more recently,
particularly owing to work in empirical systems [@Scheffer2015].
Our analysis in the origins of noise provides sufficient framework to illustrate
this idea completely.  The premise behind the tipping point just described
is that of a fold bifurucation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
The key thing we need to know about this bifurcation is that it involves
changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]$ in
the notation of our example), which can build up for some time without
significant changes to the location of the equilibrium state $\hat x$.
This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t}} = b(x) - d(x)$,
but featured regularly in our equations for the noise, such as the 
expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
This term represents the *stability*  of the equilibrium: how quickly
deviations return to the stationary value.

The trick of the early warning signals approach is to view noise as 
a sequence of countless miniature perturbation experiments.

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.

This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure \ref{tipping}.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure \ref{tipping}.

@@

<!--
One of the most interesting aspects of noise is as a source of
additional information.  This can be realized in a variety of ways.
For instance, in deriving expressions for the variance created by
demographic noise above, we observed that different partitions of
a logistic growth model can lead to the same macroscopic dynamics
(same values for $b$ - $d$), while having different variance (due
to different values for $b$+$d$).  This suggests that the variation
observed in such a population could potentially help resolve
how dynamics are partitioned between individual birth and death
rates (though in practice one would have to first account for any
contribution of environmental noise).
-->
